---
title: 'MECH481A6: Engineering Data Analysis in R'
subtitle: 'Chapter 10 Homework: Measurement' 
author: 'Brad Portouw'
date: '`r format(Sys.time(), "%d %B, %Y")`'
output: pdf_document
---

```{r global-options, include=FALSE}
# set global options for figures, code, warnings, and messages
knitr::opts_chunk$set(fig.width=6, fig.height=4, fig.path="../figs/",
                      echo=FALSE, warning=FALSE, message=FALSE)
```

# Load packages

```{r load-packages, message=FALSE}
# load packages for current session
library(tidyverse) 
library(gridExtra) 
library(MASS)
library(lubridate)
```

# Chapter 10 Homework

This homework will give you practice at working with a measurement dataset: `airlift_mass_repeatability.csv`. This data set represents repeated measures of "blank" air sampling filters. 

A couple notes to consider when reporting answers in response to questions. The microbalance used to make these measurements reads out to the nearest microgram ($\mu g$), which is 0.000001 $g$ or 0.001 $mg$. Thus, be careful when reporting descriptive statistics so as not to overstate your **precision**. Use the `round()` function to avoid reporting more than 0.1 $\mu g$ of precision (or 0.0001 $mg$). Here is some example code that uses the `across()` function from `dplyr::` to round numeric output to just four digits (appropriate for $mg$ units in this exercise):

`dplyr::mutate(across(.cols = where(is.numeric), .fns = round, 3))`

\newpage

## Question 1
Import the `airlift_mass_repeatability.csv` file into a data frame called `blanks` and perform the following data wrangling in a single pipe:  

- retain only the first 3 columns of data;
- rename the columns with the names `date`, `id`, and `mass_mg`;
- convert the `date` column vector into a date class object using `lubridate::`
- convert the `id` variable to a class `factor` (this can be accomplished using `base::as.factor()` or `purrr::as_factor()`)
- create a new column vector named `mass_mg` by rescaling the `mass_g` data (i.e., convert $g$ to $mg$ by multiplying `mass_g` by 1000)

```{r import-clean-data}
#note that the CSV data is within the Ch10_homework folder.
blanks <- read.csv('AIRLIFT_mass_repeatability.csv')%>%
  dplyr::select(1:3)%>%
  dplyr::rename(date=Date,
                id=Filter.ID,
                mass_g=Mass..g.)%>%
  dplyr::mutate(across(.cols = where(is.numeric),
                       .fns = round, 6))%>%
  mutate(date = lubridate::dmy(date),
         id = base::as.factor(id),
         mass_mg = mass_g*1000)

```

## Question 2:  

  2a. Are there any NAs present in the data frame?  
  2b. How many unique filter IDs are present in this data frame?  
  2c. How many samples are present for each filter ID? Hint: look up the `dplyr::count()` function.  
  2d. Over how long of a period were these blank measurements made? Hint: this can be done in base R with a `max() - min()` or with `lubridate::interval() %>% as.duration()`.
```{r unique-IDs}
#Taking the sum of the is.na.data.frame in order to count any na's that occur in the data frame. 
num_na <- sum(is.na.data.frame(blanks))

#
x<-unique(blanks$id)

#counting the number of samples for each filter ID
dplyr::count(blanks, id)

#Period of time the measurements were made:
meas_days <- max(blanks$date)-min(blanks$date) 


```
Seemingly there are no NA entries for the blanks data frame. The data from the raw airlift_mass_repeatability.csv did have many columns with NA's present, however those columns were removed for this data analysis.

The vector x, lists all the unique id's for the different sensors used with 5 sensors total being:

41666, 41667, 41668, 41669, 41671

. Line 73 counts the number of data points for each sensor, placing them in the data frame above. Four sensors have 78 datapoints, with sensor 41669 having 76.

The sensors measured data over the course of 35 days, as shown in the variable meas_days. 


## Question 3

Group the `blanks` data frame by `id` and calculate mean, median, and standard deviations for each filter id.  Hint: use `group_by() %>% summarise()` to do this efficiently.

```{r blank-descriprives}
by_id<- blanks %>%
  group_by(id) %>%
  summarise(mean_mass_mg = mean(mass_mg),
           median = median(mass_mg),
           SD = sd(mass_mg))
by_id
```



## Question 4

Calculate the limit of detection (LOD) for this measurement method. Note: you will need to calculate standard deviations for each filter `id` (as done in question 3) and then estimate LOD from $LOD = 3\cdot \sigma_b$ where $\sigma_b$ is calculated for each filter `id`.
```{r calculate LOD}
LOD <- by_id%>%
  dplyr::select(-median,-mean_mass_mg)%>%
  dplyr::rename(LOD = SD)%>%
  dplyr::mutate(LOD = LOD*3)

LOD
```
Above is a tibble with the limits of detection for each Sensor, all in mg. 
