---
title: 'Engineering Data analysis in R Homework 8'
subtitle: 'Chapter 8 Homework: Functional Programming' 
author: 'Brad Portouw'
date: '`r format(Sys.time(), "%d %B, %Y")`'
output: pdf_document
---

```{r global-options, include=FALSE}
# set global options for figures, code, warnings, and messages
knitr::opts_chunk$set(fig.width=6, fig.height=4, fig.path="../figs/",
                      echo=FALSE, warning=FALSE, message=FALSE)
```

# Load packages

```{r load-packages, message=FALSE}
# load packages for current session
library(tidyverse) 
library(lubridate)
library(magrittr)
library(gridExtra) # needed for extra credit question
# Regex functions struggle unless working directory is set to the original R folder. 
#Normally I work from the HW folder because that is where the Rmd is located.
# in this case I am moving up 2 directories from Ch8_homework to Homework to Portouw-MECH 476
setwd("../..")
```

# Chapter 8 Homework

This homework will give you practice at writing functions, mapping functions, and cleaning/plotting data. 

When a question asks you to make a plot, remember to set a theme, title, 
subtitle, labels, colors, etc. It is up to you how to personalize your plots, 
but put in some effort and make the plotting approach consistent throughout
the document. For example, you could use the same theme for all plots. I also
like to use the subtitle as a place for the main summary for the viewer.

\newpage

## Question 1

Write a **function** named `sort_abs()`that takes a vector of numbers as input, calculates the absolute values of each entry, and then outputs that vector sorted from smallest to largest value. 

```{r sort-baseR, include=FALSE}

#create a function named `sort_abs()`
sort_abs <- function(x, na.rm = TRUE){
  if (na.rm == FALSE){
  y <-abs(x)
  z <- sort(y)
  return(z)
  }
  else {
    y <- abs(na.omit(x))
    z <- sort(y)
    return(z)
  }
}
```

## Question 2

Modify the function `import.w.name()` to import the “date” part of the filename (in addition to the sensor ID). Create a new column variable called “date_created” with this information. Hint: you will need to apply a regex pattern like this: `"(?<=_)[:alnum:]+(?=\\.)"`

```{r import.w.name2, include=FALSE}
# create an object that tracks the file names and file paths
# see the coursebook for details
# copied from textbook ch 8.3.2
setwd("../..")
file_list <- list.files('./Data/purpleair/', full.names=TRUE)

# modify the import.w.name function
# hint: start with what is provided in the coursebook
import.w.name <- function(pathname) {
  #create a tibble by importing the 'pathname' file
  df <- read_csv(pathname, col_names = TRUE)
  df <- df %>%
    # use stringr::str_extract & a regex to get sensor ID from file name
    # regex translation: "look for a /, then extract all letters and numbers that follow until _"
    #in this case I had to change the regex to look for "purpleair/" instead of "//" because i couldn't import the path name with the list.files exactly as it was shown in the chapter 8 example. 
    mutate(sensor_ID = str_extract(pathname, 
                                  "(?<=purpleair/)[:alnum:]+(?=_)")) %>%
    #next extract the date from the file name and likely make it a date object.
    mutate(date_created =
      ymd(
        str_extract(pathname,
                                      "(?<=_)[:alnum:]+(?=\\.)")))%>%
    # return only a few salient variables to the resultant data frame using dplyr::select
    select(UTCDateTime, 
           current_temp_f, 
           current_humidity, 
           pressure,
           pm2_5_atm,
           sensor_ID,
           date_created) %>%
    na.omit() # remove NA values, which happens when sensor goes offline
  return(df)
}

PA_data_1 <- import.w.name(file_list[1])

head(PA_data_1)

```

## Question 3
This question is designed to give you practice at data cleaning. First, create a pipeline that (1) uses `purrr::map_dfr()` and `import.w.name()` to read in  all the the PurpleAir data files into a single data frame. Call that new data frame `PA_data_merged`. (2) Then, have the pipeline convert the character vector UTCDateTime into new column of class POSIXct using a `lubridate::` function (note - not all the indices in UTCDateTime will parse correctly; we will address this in Question 4). Finally, (3) finish the pipeline by renaming the `current_temp_f` and `current_humidity` column names to shorter names.

```{r map-dplyr-pipe, warning=FALSE, message=FALSE, include=FALSE}
# the map code is provided in the coursebook
#going to set wd to two directories above so function works properly.
setwd("../..")

PA_data_merged <- map_dfr(file_list,
                          import.w.name)%>%
  mutate(UTCDateTime =
    lubridate::ymd_hms(UTCDateTime))%>%
  mutate(UTCDateTime = as.POSIXct(UTCDateTime))%>%
  rename(temp_f = current_temp_f,
         humid = current_humidity)
glimpse(PA_data_merged)

```

## Question 4
Can you find the 3 indices of UTCDateTime in PA_data_merged that failed to parse with lubridate::? Hint: use the `is.na()` function nested wihtin `which()` to return the row numbers in question. Both of these are baseR functions. Once you have the row entries identified you can `View()` them with a call to `slice()`.: normal entries in UTCDateTime are all the same number of characters nchar() or entries that failed to parse in the new date column will have NA associated with them.

I'm assuming we are keeping these rows. 
```{r find-NAs, include=FALSE}
setwd("../..")

NA_date_rows <- which(is.na(PA_data_merged))

# from my understanding the question isn't asking for these rows to be removed, so I won't do that here. 

View(slice(PA_data_merged, NA_date_rows))
NA_date_rows
```
Rows 1132, 5274 and 5361 all have NA values for their UTC times.


## Question 5
Create a series of EDA plots (cdf, boxplot, histogram, time series) of the `pm2_5_atm` variable from `PA_data_merged`. Use `color =` or `fill =` as an aesthetic to differentiate each sensor by `sensor_ID`. Do the data have a central tendency? Do they appear normally distributed? Do events show up in the time series?  Note: the variable `pm2_5_atm` is the concentration of fine particulate matter air pollution in micrograms per cubic meter ($\mu g/m^3$).

```{r plots, echo=FALSE}
#going to make a cumulative fraction of the given pollution data.

pol_conc <- PA_data_merged%>%
  dplyr::group_by(sensor_ID)%>%
  dplyr::select(sensor_ID,
                 UTCDateTime,
                 pm2_5_atm)


ecdf <-ggplot(data = pol_conc,
                   mapping = aes(x=pm2_5_atm,
                       color = sensor_ID,
                       y = pm2_5_atm))+
  geom_step(stat = "ecdf")+
  theme_bw()+
  labs(title = "Cumulative Distributon of Air polluiton Concentration",
       subtitle = "Concentration of fine particulate matter in micrograms per cubic meter",
       x="Particulate matter micrograms per cubic meter",
       y="Quantile")
#Boxplot
box <- ggplot(data = pol_conc,
              mapping = aes(x= pm2_5_atm,
                            color=sensor_ID,
                            y = sensor_ID))+
  geom_boxplot(outlier.alpha = 0.4)+
  theme_bw()+
  labs(title = "Air Pollution Concentraion",
       subtitle = "Fine particulate matter in micrograms per cubic meter",
       x = "Micrograms per Cubic Meter",
       y = "Sensor ID")
#Histogram
hist <- ggplot(data=pol_conc,
               mapping = (aes(x=pm2_5_atm,
                           color = sensor_ID)))+
  geom_histogram(fill ="white",
                 position = "dodge",
                 bins=40,
                 alpha = 0.3)+
  theme_bw()+
  labs(title = "AIr Pollution Concentration",
       subtitle = "Fine particulate matter in micrograms per cubic meter",
       x = "Micrograms per Cubic Meter",
       y = "Counts")
#time series
time <- ggplot(data = pol_conc,
               mapping = aes(x=UTCDateTime,
                             color = sensor_ID,
                             y = pm2_5_atm))+
  geom_jitter(alpha = 0.3,
              size = 0.5)+
  theme_bw()+
    labs(title = "AIr Pollution Concentration",
       subtitle = "Fine particulate matter in micrograms per cubic meter at given dates",
       x = "Date",
       y = "Micrograms per Cubic Meter")

ecdf
box
hist
time
```


### Extra Credit
Create the EDA figures within a single plot (hint: use the `gridExtra::` package).  Show only one legend and place it within the body of the CDF plot (hint: to move or remove a legend, add a call that uses a version of `theme(legend.position = ...)`).

```{r PA-EDA-plots, echo=FALSE, message=FALSE, warning=FALSE}
library("gridExtra")
#cdf plot


ecdf <-ggplot(data = pol_conc,
                   mapping = aes(x=pm2_5_atm,
                       color = sensor_ID,
                       y = pm2_5_atm))+
  geom_step(stat = "ecdf")+
  theme_bw()+
  theme(legend.position = "none",
        axis.title.x = element_blank(),
        axis.title.y = element_blank())


#boxplot
box <- ggplot(data = pol_conc,
              mapping = aes(x= pm2_5_atm,
                            color=sensor_ID,
                            y = sensor_ID))+
  geom_boxplot(outlier.alpha = 0.4)+
  theme_bw()+
  theme(legend.position = "none")

#time-series plot
time <- ggplot(data = pol_conc,
               mapping = aes(x=UTCDateTime,
                             color = sensor_ID,
                             y = pm2_5_atm))+
  geom_jitter(alpha = 0.3,
              size = 0.5)+
  theme_bw()+
  theme(legend.position = "none")

#histogram
hist <- ggplot(data=pol_conc,
               mapping = (aes(x=pm2_5_atm,
                           color = sensor_ID)))+
  geom_histogram(fill ="white",
                 position = "dodge",
                 bins=40,
                 alpha = 0.3)+
  theme_bw()+
  theme(legend.position = "none",)

all_plots2 <- gridExtra::grid.arrange(ecdf,
                              box,
                              hist,
                              time,
                              widths=c(0.2,1,0.2),
layout_matrix= rbind(c(NA,1,NA),
                     c(NA,1,NA),
                     c(NA,2,NA),
                     c(NA,2,NA),
                     c(NA,3,NA),
                     c(NA,3,NA),
                     c(NA,4,NA),
                     c(NA,4,NA)))


#Combining each into one plot object:
#all_plots <- grid.arrange(ecdf,
 #                         box,
  #                        time,
   #                       hist,
    #                      layout_matrix = matrix(c(1,1,2,2,2,3,3,4,4,NA),
                                                 #byrow = TRUE,
  #                        ncol=5))
#all_plots
```
